---
title: "RCT estimator"
author: Bénédicte Colnet
output: html_document
---

This notebook gathers functions to estimate the average treatment effect from a RCT sample with the difference in means estimator, but also the conditional OLS estimator. This is to illustrate the good properties of the two estimators, and to show that the OLS estimator has a lower variance even if the model is badly specified.

In this notebook simulations are performed, and in the linear and non linear case, the ATE is of 10.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(MASS) # simulation
library(ggplot2)
```

```{r}
difference_in_means <- function(dataset) {
  treated_idx <- which(dataset$W == 1)
  control_idx <- which(dataset$W == 0)
  
  # Filter treatment / control observations, pulls outcome variable as a vector
  y1 <- dataset[treated_idx, "Y"] # Outcome in treatment grp
  y0 <- dataset[control_idx, "Y"] # Outcome in control group
  
  n1 <- sum(dataset$W)     # Number of obs in treatment
  n0 <- sum(1 - dataset$W) # Number of obs in control

  # Difference in means is ATE
  tauhat <- mean(y1) - mean(y0)
  
  # 95% Confidence intervals
  se_hat <- sqrt( var(y0)/(n0-1) + var(y1)/(n1-1) )
  lower_ci <- tauhat - 1.96 * se_hat
  upper_ci <- tauhat + 1.96 * se_hat
  
  return(c(ATE = tauhat, lower_ci = lower_ci, upper_ci = upper_ci))
}
# note that t.test on the two groups give you the same or a regression


difference_in_condmean_ols <- function(dataset) {
  n <- dim(dataset)[1]
  treated_idx <- which(dataset$W == 1)
  control_idx <- which(dataset$W == 0)
  
  # Filter treatment / control observations, pulls outcome variable as a vector
  y1 <- dataset[treated_idx, "Y"] # Outcome in treatment grp
  y0 <- dataset[control_idx, "Y"] # Outcome in control group
  
  n1 <- sum(dataset$W)     # Number of obs in treatment
  n0 <- sum(1 -dataset$W) # Number of obs in control
  
  #dataset_spread <- fastDummies::dummy_cols(dataset, remove_selected_columns = T)
  dataset_spread <- dataset
  df0 <- dplyr::select(dataset_spread[control_idx,], -c("W"))
  df1 <- dplyr::select(dataset_spread[treated_idx,], -c("W"))
  
  mu0 <- lm(Y ~ ., data = df0)
  mu1 <- lm(Y ~ ., data = df1)
  
  # Difference in predicted means is ATE
  tauhat <- mean(predict(mu1, newdata = dplyr::select(dataset_spread, -c("W","Y"))) - predict(mu0, newdata = dplyr::select(dataset_spread, -c("W","Y"))))
  
  # 95% Confidence intervals
  beta0 <- mu0$coefficients[-1]
  beta0[which(is.na(beta0))] <- 0
  beta1 <- mu1$coefficients[-1]
  beta1[which(is.na(beta1))] <- 0
  se_hat <- sqrt( n*(var(y0)/(n0) + var(y1)/(n1)) - t(beta0+beta1)%*%var(dplyr::select(dataset_spread, -c("W","Y")))%*%(beta0+beta1))/sqrt(n)
  lower_ci <- tauhat - 1.96 * se_hat
  upper_ci <- tauhat + 1.96 * se_hat
  
  return(c(ATE = tauhat, lower_ci = lower_ci, upper_ci = upper_ci))
}

```


```{r}
linear_simulation <- function(samples = 5000){
  X = mvrnorm(n=samples, mu=c(1, 1, 1, 1), Sigma=diag(4))
  data <- as.data.frame(X)
  names(data) <- paste("X", 1:4, sep = "")
  data$W <- rbinom(nrow(data), 1, 0.5)
  
  # generate linear outcome
  error = rnorm(n = samples, mean = 0, sd = 0.2)
  data$Y =3*data$X1 + 2*data$X2 + - 2*data$X3 - 0.8*data$X4 + data$W*(2*data$X1 +5*data$X3 + 3*data$X4) + error
  
  return(data)
}

non_linear_simulation <- function(samples = 5000){
  X = mvrnorm(n=samples, mu=c(1, 1, 1, 1), Sigma=diag(4))
  data <- as.data.frame(X)
  names(data) <- paste("X", 1:4, sep = "")
  data$W <- rbinom(nrow(data), 1, 0.5)
  
  # generate non linear outcome
  error = rnorm(n = samples, mean = 0, sd = 0.2)
  data$Y = 3*data$X1 + 2*data$X2*data$X2 - 2*data$X3*0.8*data$X4 + data$W*10 + error
  
  return(data)
}
```


```{r}
results <- data.frame("estimator" = c(),
                             "estimate" = c(),
                             "linear" = c())

for (i in 1:100){
  sim_lin <- linear_simulation()
  tau_dm <- difference_in_means(sim_lin)
  new_result_lin = data.frame("estimator" = "DM", "estimate" = tau_dm["ATE"], "linear" = TRUE)
  tau_ols <- difference_in_condmean_ols(sim_lin)
  new_result_lin_ols = data.frame("estimator" = "OLS", "estimate" = tau_ols["ATE"], "linear" = TRUE)
  results <- rbind(results, new_result_lin, new_result_lin_ols)

  sim_nlin <- non_linear_simulation()
  tau_dm <- difference_in_means(sim_nlin)
  new_result_nlin = data.frame("estimator" = "DM", "estimate" = tau_dm["ATE"], "linear" = FALSE)
  tau_ols <- difference_in_condmean_ols(sim_nlin)
  new_result_nlin_ols = data.frame("estimator" = "OLS", "estimate" = tau_ols["ATE"], "linear" = FALSE)
  results <- rbind(results, new_result_nlin, new_result_nlin_ols)
}
```



```{r}
library(ggplot2)
ggplot(results, aes(x = estimator, y = estimate)) +
  geom_boxplot() +
  facet_grid(~linear) +
  theme_bw()
```


